---
title: WoodLabFunctions
output:
  md_document:
    variant: markdown_github
---

# WoodLabFunctions by afpybus

A collection of commonly used functions by the grad students, lab techs, and undergrads of the Wood Lab.

## Installation

Copy and paste the following code into your R console:

```{r Installation, eval = FALSE}
# Install
if(!require(devtools)) install.packages("devtools")
devtools::install_github("afpybus/WoodLabFunctions")
```

## Multivariate Analysis

The Wood Lab routinely conducts multivariate analysis in R, including principal component analysis (PCA), partial least squares regression analysis (PLSR), and partial least squares discriminant analysis (PLS-DA). We refer the reader to *Multi- and Megavariate Data Analysis Basic Principles and Applications* by L. Eriksson for an in-depth discussion of each.

Load a sample data set from our first repetitive mild traumatic brain injury (rmTBI) study:

```{r Load Data}
library("WoodLabFunctions")
data(rmTBI)

# Display the first 6 rows and 8 columns
head(rmTBI[,1:8],6)
```

This data set is derived from eight mice (8 rows) that received three rmTBI. We measured 32 different proteins (32 columns) after injury and were interested in whether one of the proteins, Iba1 (a microglial activation marker), correlated with the rest of the measured proteins. The other 31 proteins are cytokines, immune signaling proteins.

### Principal Component Analysis (PCA)

Partial least squares regression using Iba1 as a predictor variable and the cytokines as response variables is an appropriate method for exploring possible relationships. We will use the ropls package to first visualize variance in the data using a PCA then to regress the data against Iba1.

```{r Run PCA}
predictor = rmTBI$Iba1
cytokines = rmTBI[,-1]

# Run the PCA
PCA = ropls::opls(cytokines,predI=2) # set predI=2 to extract just the first two components
```

Next, use rotate_opls() to create output that's compatible for scores_plot() and loadings_chart().

```{r First Scores Plot, fig.width=4, fig.height=3.5, warning=FALSE}
PCA_rot = rotate_opls(PCA)
scores_plot(T1=PCA_rot$T1,T2=PCA_rot$T2,color = rownames(cytokines))
```

Let's now color the data points by their Iba1 value to get a sense for the distribution. We can additionally specify that the analysis was a PCA to update the axis labels.

```{r PCA Gradient, fig.width=4, fig.height=3.5, warning=FALSE}
# Use scores_plot_gradient instead of scores_plot
scores_plot_gradient(T1=PCA_rot$T1,T2=PCA_rot$T2, 
                     color = predictor, # Set the colors to the Iba1 values stored in predictor
                     analysis = "PCA") # Updates plot title and changes axis labels to "PC1" and "PC2"
```

### Partial Least Squares Regression (PLSR)

Next, we'll conduct PLSR to create latent variables associated with changes in Iba1. We'll use opls() again:

```{r Run PLSR}
predictor = rmTBI$Iba1
cytokines = rmTBI[,-1]

# Run the PLSR
PLSR = ropls::opls(cytokines,
                   y=predictor, # set a y variable to change from PCA to PLS
                   predI=2) # set predI=2 to extract just the first two components
```

```{r PLSR Gradient, fig.width=4, fig.height=3.5, warning=FALSE}
PLSR_rot = rotate_opls(PLSR)
scores_plot_gradient(T1=PLSR_rot$T1,T2=PLSR_rot$T2, 
                     color = predictor, # Set the colors to the Iba1 values stored in predictor
                     color.str = "Iba1",
                     analysis = "PLS") # Updates plot title and changes axis labels to "PC1" and "PC2"

```

Let's maximize variance across Iba1 in LV1 by conducting a small rotation of the model.

```{r PLSR Rotated, fig.width=4.3, fig.height=3.5, warning=FALSE}
PLSR_rot30 = rotate_opls(PLSR,degrees=30)
scores_plot_gradient(T1=PLSR_rot30$T1,T2=PLSR_rot30$T2, 
                     color = predictor, # Set the colors to the Iba1 values stored in predictor
                     color.str = "Iba1",
                     analysis = "PLS") # Updates plot title and changes axis labels to "PC1" and "PC2"

```

Now that we've roughly aligned our greatest variance in Iba1 with LV1, let's figure out what drives LV1. We will construct a loadings chart which will indicate the contribution of each cytokine to the sample scores along LV1. Highly positive "loadings" indicate correlation with Iba1 (positive LV1 scores), while negative loadings are associated with negative LV1 scores.

```{r First Loading Chart, fig.width=8, fig.height=3.2, warning=FALSE}
loadings_chart(loadings=PLSR_rot30$P1, 
               component_str = "LV1")
```

### PLSR with Leave One Out Cross Validation

To ensure the results of the model are not heavilty dependent on any one sample, we will conduct a leave one out cross validation with 25 iterations. We'll then add error bars depicting the standard deviation across all iterations.

```{r PLSR with LOOCV}
PLSR_LOOCV = opls_LOOCV(cytokines,
                   y=predictor, 
                   runs=25) # sets number of iterations
```

```{r PLSR LOOCV Loadings, fig.width=8, fig.height=3.2, warning=FALSE}
PLSR_LOOCV_rot30 = rotate_opls_LOOCV(PLSR_LOOCV,degrees=30)
loadings_chart_sd(loadings=PLSR_LOOCV_rot30$P1, 
                  sd = PLSR_LOOCV_rot30$sd_P1_LOOCV,
                  component_str = "LV1",
                  lim = 1.1) # set a higher axis limit to see full error bars
```
