---
title: WoodLabFunctions
output:
  md_document:
    variant: markdown_github
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-",
  warning = FALSE
)
```

# WoodLabFunctions by afpybus

A repository of commonly used functions by the grad students, lab techs, and undergrads of the Wood Lab.

## Installation

Copy and paste the following code into your R console:

```{r Installation, eval = FALSE}
# Install
if(!require(devtools)) install.packages("devtools")
devtools::install_github("afpybus/WoodLabFunctions")
```

## Multivariate Analysis

The Wood Lab routinely conducts multivariate analysis in R, including principal component analysis (PCA), partial least squares regression analysis (PLSR), and partial least squares discriminant analysis (PLS-DA). We refer the reader to *Multi- and Megavariate Data Analysis Basic Principles and Applications* by L. Eriksson for an in-depth discussion of each.

Load a sample data set from our first repetitive mild traumatic brain injury (rmTBI) study:

```{r Load Data}
library("WoodLabFunctions")
data(rmTBI)

# Display the first 6 rows and 8 columns
head(rmTBI[,1:8],6)
```

This data set is derived from eight mice (8 rows) that received three rmTBI. We measured 32 different proteins (32 columns) after injury and were interested in whether one of the proteins, Iba1 (a microglial activation marker), correlated with the rest of the measured proteins. The other 31 proteins are cytokines, immune signaling proteins. Let's take our first look at the data by creating a heatmap:

```{r Heatmap, fig.width=6, fig.height=4, warning=FALSE}
hm = heatmap_wl(rmTBI,
                mar=c(10,2),  # set margins to re-size the heatmap
                clust_r = TRUE, # cluster rows by euclidean distance
                clust_c = TRUE, # cluster cols by euclidean distance
                labRow=NA, # remove row labels
                cex_c=0.5) # re-size col labels to fit all of them
```

### Principal Component Analysis (PCA)

Partial least squares regression using Iba1 as a predictor variable and the cytokines as response variables is an appropriate method for exploring possible relationships. We will use the ropls package to first visualize variance in the data using a PCA then to regress the data against Iba1 using PLSR.

```{r Run PCA}
predictor = rmTBI$Iba1
cytokines = rmTBI[,-1]

# Run the PCA
PCA = ropls::opls(cytokines,predI=2) # set predI=2 to extract just the first two components
```

Next, use rotate_opls() to create output that's compatible for scores_plot() and loadings_chart().

```{r First Scores Plot, fig.width=4, fig.height=3.5, warning=FALSE}
PCA_rot = rotate_opls(PCA)
scores_plot(T1=PCA_rot$T1,T2=PCA_rot$T2,color = rownames(cytokines))
```

Let's now color the data points by their Iba1 value to get a sense for the distribution. We can additionally specify that the analysis was a PCA to update the axis labels.

```{r PCA Gradient, fig.width=4, fig.height=3.5, warning=FALSE}
# Use scores_plot_gradient instead of scores_plot
scores_plot_gradient(T1=PCA_rot$T1,T2=PCA_rot$T2, 
                     color = predictor, # Set the colors to the Iba1 values stored in predictor
                     analysis = "PCA") # Updates plot title and changes axis labels to "PC1" and "PC2"
```

### Partial Least Squares Regression (PLSR)

Next, we'll conduct PLSR to create latent variables associated with changes in Iba1. We'll use opls() again:

```{r Run PLSR}
predictor = rmTBI$Iba1
cytokines = rmTBI[,-1]

# Run the PLSR
PLSR = ropls::opls(cytokines,
                   y=predictor, # set a y variable to change from PCA to PLS
                   predI=2) # set predI=2 to extract just the first two components
```

```{r PLSR Gradient, fig.width=4, fig.height=3.5, warning=FALSE}
PLSR_rot = rotate_opls(PLSR)
scores_plot_gradient(T1=PLSR_rot$T1,T2=PLSR_rot$T2, 
                     color = predictor, # Set the colors to the Iba1 values stored in predictor
                     color.str = "Iba1",
                     analysis = "PLS") # Updates plot title and changes axis labels to "PC1" and "PC2"

```

Let's maximize variance across Iba1 in LV1 by conducting a small rotation of the model.

```{r PLSR Rotated, fig.width=4.3, fig.height=3.5, warning=FALSE}
PLSR_rot30 = rotate_opls(PLSR,degrees=30)
scores_plot_gradient(T1=PLSR_rot30$T1,T2=PLSR_rot30$T2, 
                     color = predictor, # Set the colors to the Iba1 values stored in predictor
                     color.str = "Iba1",
                     analysis = "PLS") # Updates plot title and changes axis labels to "PC1" and "PC2"

```

Now that we've roughly aligned our greatest variance in Iba1 with LV1, let's figure out what drives LV1. We will construct a loadings chart which will indicate the contribution of each cytokine to the sample scores along LV1. Highly positive "loadings" indicate correlation with Iba1 (positive LV1 scores), while negative loadings are associated with negative LV1 scores.

```{r First Loading Chart, fig.width=8, fig.height=3.2, warning=FALSE}
loadings_chart(loadings=PLSR_rot30$P1, 
               component_str = "LV1")
```

### PLSR with Leave One Out Cross Validation

To ensure the results of the model are not heavily dependent on any one sample, we will conduct a leave one out cross validation with 25 iterations. We'll then add error bars depicting the standard deviation across all iterations.

```{r PLSR with LOOCV}
PLSR_LOOCV = opls_LOOCV(cytokines,
                   y=predictor, 
                   runs=25) # sets number of iterations
```

```{r PLSR LOOCV Loadings, fig.width=8, fig.height=3.3, warning=FALSE}
PLSR_LOOCV_rot30 = rotate_opls_LOOCV(PLSR_LOOCV,degrees=30)
loadings_chart_sd(loadings=PLSR_LOOCV_rot30$P1, 
                  sd = PLSR_LOOCV_rot30$sd_P1_LOOCV,
                  component_str = "LV1",
                  lim = 1.2) # set a higher axis limit to see full error bars
```

Finally, we can assess the performance of the PLSR by computing a regression of Iba1 with scores on LV1. We can show this graphically using scatter_reg():

```{r Scatter_reg,fig.height=4,fig.width=4}
scatter_reg(x=PLSR_LOOCV_rot30$T1,y=predictor,
            xlab = "Scores on LV1",
            ylab = "Iba1 [a.u.]",
            title = "LV1 vs Iba1")
```

## Linear Range Analysis

Assessing the linear range of detection for multiplex ELISA kits is a key quality control step that must be completed before sample measurement. In short, the user selects a single sample and loads increasing amounts of lysate into multiple assay wells. After completion of the ELISA, the expected result is that increased total protein loading correlates with increased detection of each protein. For proteins that do not show this linear increase in detection with increases in loading, they are considered to be outside of the detectable linear range for the chosen loadings.

Let's load some example linear range data to explore. The first column contains values for mass of total protein [ug] and each subsequent column contains detected fluorescent values [a.u.] of a phospho-protein from our MAPK Luminex assay.

```{r Load LR Data}
data(LinearRange)

LinearRange
```

First, let's construct an x-y scatter graph for a single measured phospho-protein to see the relationship between detected analyte and loaded total protein using scatter_reg():

```{r Scatter Graph, fig.width=4,fig.height=4}
scatter_reg(x=LinearRange$Mass,y=LinearRange$pJnk,
            xlab="Protein Mass [ug]",ylab="pJnk [a.u.]",
            showRSQ=FALSE,showP=FALSE,
            method="loess")

```

The most important outcome from your linear range will be the amount of protein to load in your subsequent assays. You will want to look at the linear range of ALL measured proteins and select the loading which is appropriate for the highest number of proteins of interest. We can do this easily using LinearRangePanel().

```{r Linear Range Panel, fig.height=10,fig.width=12}
loadings = LinearRange$Mass
analytes = LinearRange[,2:10]

panel = LinearRangePanel(analytes = analytes,loadings = loadings,xlab = "Protein Mass [ug]")

```

We selected a loading of 1ug, which we can display on the graphs to help assess its suitability:

```{r Linear Range Panel vline, fig.height=10,fig.width=12}
loadings = LinearRange$Mass
analytes = LinearRange[,2:10]

panel = LinearRangePanel(analytes = analytes,loadings = loadings,xlab = "Protein Mass [ug]",xintercept = 1)

```

In this example, we would have high confidence in measured values for pAtf2, pMek1, pJnk, pErk1/2, pHSP27, and pStat3. p-p38 is outside of its detectable linear range.
